---
title: "Could anyone have been harmed?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Could anyone have been harmed?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CMRSS)
# Few simulated assignments and a coarse precision keep the vignette quick to
# build; use nperm = 10^4 or more and the default tol in a real analysis.
nperm <- 1000
tol <- 0.05
```

A program evaluation reports that a job-training program raised average
earnings. A city council member asks a different question: did the program
make anyone worse off? The average cannot answer that. An average gain of
\$500 is what we would see if every participant gained \$500, and it is also
what we would see if most participants gained \$1500 while some lost \$1500.

This vignette shows how to use `cmrss()` to answer the council member's
question with data from a randomized experiment. The answer takes the form
"with 95 percent confidence, at least L units were harmed", and, alongside
it, "at most U units could have been harmed".

## What we can and cannot learn

Each unit i has two potential outcomes: $Y_i(1)$, the outcome it would have
if treated, and $Y_i(0)$, the outcome it would have if not. Its individual
effect is $\tau_i = Y_i(1) - Y_i(0)$. A unit was harmed if $\tau_i < 0$. We
never see both potential outcomes for the same unit, so we never see any
$\tau_i$, and we cannot say which units were harmed.

We can still learn how many. Sort the n individual effects from smallest to
largest, $\tau_{(1)} \le \tau_{(2)} \le \dots \le \tau_{(n)}$. Randomization
lets us test hypotheses such as "$\tau_{(k)} \le c$": the k-th smallest
effect is at most c. These tests are one-sided: a large c is never
rejected, and a small c is rejected when the data show $\tau_{(k)}$ must be
larger. The smallest c that the test does not reject is a lower confidence
bound for $\tau_{(k)}$. If the lower
bound for $\tau_{(k)}$ is above 0, then $\tau_{(k)}$ and every larger effect,
$\tau_{(k)}, \dots, \tau_{(n)}$, are above 0, so at least $n - k + 1$ units
have positive effects. `cmrss()` computes these bounds for every k and does
the counting.

## Turning lower bounds into a count of harmed units

The package computes lower bounds only. A unit was harmed if its effect is
below 0, and a lower bound cannot show that an effect is below anything.
The harm question needs upper bounds, and we get them by analyzing minus
the outcome. Each unit's effect on $-Y$ is
$(-Y_i(1)) - (-Y_i(0)) = -\tau_i$. A unit harmed on Y, with $\tau_i < 0$,
has a positive effect on $-Y$. So a count of units with effects on $-Y$
above 0 is a count of units harmed on Y. In `cmrss()` this is a minus sign
in the formula: `cmrss(-y ~ z, data = d)`.

If larger outcomes are worse, as with blood pressure or days absent, harm
means $\tau_i > 0$, and the analysis of y itself counts the harmed units.

## An example in which some units are harmed

We simulate an experiment in which we know the truth. Of 200 units, 100 gain
2 points, 40 lose 3 points, and 60 are unaffected. Half the units are
assigned to treatment at random.

```{r simulate}
set.seed(20261004)
n <- 200
y0 <- rnorm(n)
tau <- rep(0, n)
who <- sample(n)
tau[who[1:40]] <- -3     # harmed
tau[who[41:140]] <- 2    # helped
z <- sample(rep(c(1, 0), each = n / 2))
sim <- data.frame(z = z, y = ifelse(z == 1, y0 + tau, y0))
c(true_average_effect = mean(tau),
  difference_in_means = mean(sim$y[sim$z == 1]) - mean(sim$y[sim$z == 0]))
```

The average effect is $(100 \times 2 - 40 \times 3) / 200 = 0.4$, and the
difference in means estimates it. Nothing in that number hints that 40 units
were harmed.

We ask two questions of the same data: how many units were helped, and how
many were harmed. Reporting both answers as one statement at 95 percent
confidence requires each analysis to run at 97.5 percent, `alpha = 0.025`,
because the chance that at least one of two statements fails is at most the
sum of their separate chances, $0.025 + 0.025 = 0.05$.

```{r sim-fit}
helped <- cmrss(y ~ z, data = sim, set = "all", alpha = 0.025,
                nperm = nperm, tol = tol)
harmed <- cmrss(-y ~ z, data = sim, set = "all", alpha = 0.025,
                nperm = nperm, tol = tol)
helped
harmed
```

`set = "all"` bounds the effects of all 200 units, treated and control. The
analysis of `-y` finds at least `r sum(harmed$bounds$lower > 0)` units with
effects on $-y$ above 0, that is, at least
`r sum(harmed$bounds$lower > 0)` harmed units. The analysis of `y` finds at
least `r sum(helped$bounds$lower > 0)` helped units. A helped unit was not
harmed, so at most
$200 - `r sum(helped$bounds$lower > 0)` =
`r 200 - sum(helped$bounds$lower > 0)`$ units could have been harmed. This
upper count includes the units with no effect, which the data cannot
separate from the harmed ones.
Together, with 95 percent confidence, the number harmed is between
`r sum(harmed$bounds$lower > 0)` and `r 200 - sum(helped$bounds$lower > 0)`.
The truth, 40, lies inside that range.

The range runs over most of the 200 units. Data from one experiment cannot
fix how many units were harmed, because the treated and control outcomes of each unit are never
seen together. The lower end is still informative: it shows that some units
were made worse off, which the average effect concealed.

The bounds behind the counts are in `harmed$bounds`, one row per sorted
effect. The first row with a lower bound above 0 marks where the counting
starts:

```{r sim-bounds}
b <- harmed$bounds
b[b$lower > 0, ][1, ]
```

## The teacher-training experiment

`electric_teachers` holds an experiment in which 164 of 233 teachers, in 7
sites, were assigned to a training program, and the outcome `gain` is the
change in a test score. We first ignore the sites and treat the experiment
as completely randomized.

```{r teachers-cre}
data(electric_teachers)
t_helped <- cmrss(gain ~ TxAny, data = electric_teachers, set = "all",
                  alpha = 0.025, nperm = nperm, tol = tol)
t_harmed <- cmrss(-gain ~ TxAny, data = electric_teachers, set = "all",
                  alpha = 0.025, nperm = nperm, tol = tol)
t_helped
t_harmed
```

Both calls warn that 29 of the 233 teachers share one value of `gain`. The
package breaks ties between outcomes by the order of the rows in the data,
so with this many ties, reordering the rows can move the bounds.

With 95 percent confidence, at least `r sum(t_helped$bounds$lower > 0)`
teachers gained from the program, and no teacher is shown to have been
harmed. That is not evidence that nobody was harmed. The same analysis
allows up to $233 - `r sum(t_helped$bounds$lower > 0)` =
`r 233 - sum(t_helped$bounds$lower > 0)`$ teachers to have been harmed. The
data say the program helped many teachers and cannot rule out that it hurt
some.

The experiment randomized teachers within sites. An analysis that uses the
sites compares teachers only with others in their own site, as the
randomization did. That requires an optimization solver, such as the
`highs` package.

```{r teachers-sre, eval = requireNamespace("highs", quietly = TRUE)}
s_harmed <- cmrss(-gain ~ TxAny | Site, data = electric_teachers,
                  set = "treat", alpha = 0.05, nperm = nperm,
                  opt.method = "ILP_highs")
s_harmed
```

`set = "treat"` bounds only the effects of the 164 treated teachers. It
answers "how many of the teachers who received the program were harmed by
it", which is often the question a council member means.

## Choices that change the answer

`set` decides whose effects are counted. `"treat"` and `"control"` give
bounds for one group at the full `alpha`. `"all"` covers every unit and
splits `alpha` between the treated and control groups.

`alpha` decides the confidence level. When the helped and harmed analyses
are reported together, run each at `alpha / 2`.

`s` and `scores` decide which rank statistics are combined. A small s
detects effects shared by most units and a large s detects large effects in
a few units. The default combines three values from 2 to a maximum set by
the number of treated units; the section "Choosing s" of `?cmrss` gives
the rule and the evidence for it.

The outcome must take many values. With a binary outcome, every individual
effect is $-1$, $0$, or $1$, the tie problem above is at its worst, and
`cmrss()` warns. For a binary outcome, use a method that works with the
2 x 2 table of potential outcomes directly, such as those of Rigdon and
Hudgens (2015, *Statistics in Medicine* 34, doi:10.1002/sim.6384) or Li and
Ding (2016, *Statistics in Medicine* 35, doi:10.1002/sim.6924).
