---
title: "Could anyone have been harmed?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Could anyone have been harmed?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CMRSS)
nperm <- 1000
tol <- 0.05
# The teacher example respects the sites, which needs an optimization solver.
has_highs <- requireNamespace("highs", quietly = TRUE)
```

A program evaluation reports that a job-training program raised average
earnings by \$500. A city council member asks a different question: did the
program make anyone worse off? The average cannot answer that question. We would see
an average gain of \$500 if every participant gained \$500. We would see the
same average if two thirds of the participants gained \$1500 and one third
lost \$1500.

With data from a randomized experiment, `cmrss()` gives the council member
two numbers, both holding together with 95 percent confidence: at least
this many participants were made worse off, and at most this many could
have been. This vignette shows how to
compute both and how to read them.

## What a randomized experiment can tell us about individual effects

Call each participant a unit. Each unit i has two potential outcomes:
$Y_i(1)$, the outcome it would have if treated, and $Y_i(0)$, the outcome it
would have if not treated. Its individual effect is
$\tau_i = Y_i(1) - Y_i(0)$. The unit was harmed if $\tau_i < 0$. We see
only one of the two potential outcomes for each unit. So we never see any
$\tau_i$. We cannot say which units were harmed.

We can still learn how many units were harmed. Sort the n individual
effects from smallest to largest, $\tau_{(1)} \le \tau_{(2)} \le \dots \le
\tau_{(n)}$. Because treatment was assigned at random, we can test the
hypothesis that the k-th smallest effect is at most some number c, written
$\tau_{(k)} \le c$. We reject that hypothesis when the observed data
would be unlikely, under random assignment, if $\tau_{(k)}$ were at most c.
The largest c that we reject at the 5 percent level is a 95 percent lower
confidence bound for $\tau_{(k)}$.

`cmrss()` returns a lower bound for every k from 1 to n. The bounds are
built so that all n of them hold together with 95 percent confidence (see
`?com_conf_quant_larger_cre`). We may therefore look through the whole list and pick out the
bounds above 0. Suppose the lower bound for $\tau_{(k)}$ is above 0. The
effects $\tau_{(k+1)}, \dots, \tau_{(n)}$ are at least as large as
$\tau_{(k)}$. So, with 95 percent confidence, all $n - k + 1$ effects from
$\tau_{(k)}$ to $\tau_{(n)}$ are above 0. At least $n - k + 1$ units have
positive effects. `cmrss()` reports this count for the smallest such k.

## Counting harmed units with a minus sign

Lower bounds count units with positive effects. To count harmed units we
need to show that some effects are below 0. A lower bound cannot show
that. To count them, we multiply every outcome by -1 and analyze the new
outcome, $-Y$, in place of $Y$. Each unit's effect on
$-Y$ is $(-Y_i(1)) - (-Y_i(0)) = -\tau_i$. A harmed unit, with
$\tau_i < 0$, therefore has a positive effect on $-Y$. The count of units
with positive effects on $-Y$ is a count of harmed units. In `cmrss()` the
change is a minus sign in the formula: `cmrss(-y ~ z, data = d)`.

When larger outcomes are worse, as with blood pressure or days absent, harm
means $\tau_i > 0$. The analysis of y itself then counts harmed units.

## A simulated program that harms some participants

We simulate an experiment in which we set every unit's effect, so we can
compare the answers with the truth. Of 200 units, 100 gain 2 points, 40
lose 3 points, and 60 are unaffected. Half the units are assigned to
treatment at random.

```{r simulate}
set.seed(20261004)
n <- 200
y0 <- rnorm(n)
tau <- rep(0, n)
who <- sample(n)
tau[who[1:40]] <- -3     # harmed
tau[who[41:140]] <- 2    # helped
z <- sample(rep(c(1, 0), each = n / 2))
sim <- data.frame(z = z, y = ifelse(z == 1, y0 + tau, y0))
diff_means <- mean(sim$y[sim$z == 1]) - mean(sim$y[sim$z == 0])
```

The average effect is $(100 \times 2 - 40 \times 3) / 200 = 0.4$ points. In
this simulated experiment the treated group's mean outcome exceeds the
control group's by `r round(diff_means, 2)` points. A researcher who reported
only that difference would tell the council member the program helped. The
difference says nothing about the 40 units the program harmed.

We ask the data two questions: how many units were helped, and how many
were harmed. We want both answers to hold together with 95 percent
confidence. The chance that at least one of two statements is wrong is at
most the sum of the two chances. So if we compute each at 97.5 percent
confidence, both hold together with at least 95 percent confidence. In `cmrss()` that is `alpha = 0.025`, where `alpha` is one minus
the confidence level. To keep this page quick to build, each call
below uses 1000 simulated random assignments (`nperm`) and finds each bound
to within 0.05 (`tol`). A real analysis should use the defaults of 10,000
assignments and 0.01.

```{r sim-fit}
helped <- cmrss(y ~ z, data = sim, set = "all", alpha = 0.025,
                nperm = nperm, tol = tol)
harmed <- cmrss(-y ~ z, data = sim, set = "all", alpha = 0.025,
                nperm = nperm, tol = tol)
helped
harmed
n_harmed <- sum(harmed$bounds$lower > 0)
n_helped <- sum(helped$bounds$lower > 0)
```

With `set = "all"`, `cmrss()` bounds the effects of all 200 units, treated
and control. It does so by bounding the treated units and the control units
separately, each at half of `alpha`. The other choices, `set = "treat"` and
`set = "control"`, bound one group at the full `alpha`. Analyzing `-y`, we find `r n_harmed` units with positive
effects on $-y$, so at least `r n_harmed` units were harmed. Analyzing `y`,
we find at least `r n_helped` helped units. A helped unit was not harmed,
so at most $200 - `r n_helped` = `r 200 - n_helped`$ units could have been
harmed. These `r 200 - n_helped` units are every unit not shown to be
helped. They include the harmed units, the unaffected units, and helped
units that the data could not detect.

With 95 percent confidence, then, the number of harmed units is between
`r n_harmed` and `r 200 - n_helped`. The true number, 40, lies in that range.
The range spans most of the 200 units, because we never see a unit's
treated and control outcomes together. The lower end still answers the
council member. At least `r n_harmed` units were made worse off, although
the average effect was positive.

Each row of `harmed$bounds` is one effect on $-y$. Its column `k` gives the
effect's position in the sorted list. Its column `lower` gives the effect's
lower bound. The number of harmed
units comes from the first row whose lower bound is above 0:

```{r sim-bounds}
b <- harmed$bounds
first <- b[b$lower > 0, ][1, ]
first
```

There k = `r first$k`, so the count is
$200 - `r first$k` + 1 = `r 200 - first$k + 1`$ harmed units.

### Every sorted effect at once

The two counts come from two lists of 200 bounds. A plot of both lists
shows the bounds for every sorted effect, not only whether each is above or
below 0. For the k-th smallest effect on $y$, $\tau_{(k)}$, the analysis of
`y` gives a lower bound directly: `helped$bounds$lower[k]`. The analysis of
`-y` gives an upper bound. Its j-th smallest effect is
$-\tau_{(n - j + 1)}$, so its lower bound for that effect, with the sign
changed, is an upper bound for $\tau_{(n - j + 1)}$. Setting
$j = n - k + 1$ gives the upper bound for $\tau_{(k)}$:
`-harmed$bounds$lower[n - k + 1]`. Because each analysis ran at 97.5
percent confidence, all 200 lower bounds and all 200 upper bounds hold
together with 95 percent confidence.

The function below draws the bounds in the layout of the figures in the
paper behind this package: the sorted position k runs up the vertical
axis, and the bound for $\tau_{(k)}$ runs along the horizontal axis.
Between the two lines lie the values of $\tau_{(k)}$ that the data do not
rule out. Where a bound is $-\infty$ or $+\infty$, meaning the data put no
limit on that side, its line runs to the edge of the box around the plot.

```{r plot-fn}
plot_bounds <- function(lower, upper, truth = NULL, xlim = NULL,
                        xlab = "individual effect") {
  n <- length(lower)
  finite <- c(lower[is.finite(lower)], upper[is.finite(upper)], truth)
  if (is.null(xlim)) xlim <- range(finite)
  plot(NA, xlim = xlim, ylim = c(1, n), xlab = xlab,
       ylab = "k, position in the sorted effects")
  # An infinite bound runs to the frame, beyond every finite bound, so it
  # cannot be mistaken for the smallest or largest finite one.
  edge <- par("usr")[1:2]
  lo <- ifelse(is.finite(lower), lower, edge[1])
  hi <- ifelse(is.finite(upper), upper, edge[2])
  segments(lo, 1:n, hi, 1:n, col = "grey85")
  lines(lo, 1:n, type = "s", lwd = 2)
  lines(hi, 1:n, type = "s", lwd = 2, lty = 2)
  if (!is.null(truth)) points(sort(truth), 1:n, pch = 16, cex = 0.4,
                              col = "red")
  abline(v = 0, col = "grey40")
}
```

```{r sim-plot, fig.width = 6, fig.height = 5, fig.alt = "Lower and upper bounds for each of the 200 sorted effects in the simulation, with the true sorted effects as points."}
sim_lower <- helped$bounds$lower
sim_upper <- -rev(harmed$bounds$lower)
plot_bounds(sim_lower, sim_upper, truth = tau)
legend("bottomright", bty = "n", bg = "white",
       legend = c("lower bound", "upper bound", "true effect"),
       lty = c(1, 2, NA), lwd = c(2, 2, NA), pch = c(NA, NA, 16),
       col = c("black", "black", "red"))
```

`rev()` puts the bounds from the analysis of `-y` in the order of k: its
last row, j = 200, bounds $\tau_{(1)}$, and its first row bounds
$\tau_{(200)}$.

The red points are the true sorted effects: 40 at $-3$, 60 at 0, and 100 at
2. Every point lies between its two lines. With 95 percent confidence
we expect this in at least 95 of every 100 experiments like this one, and it
happened in this one. The top `r sum(sim_lower > 0)` rows have a lower bound above 0;
those are the units counted above as helped. The bottom
`r sum(sim_upper < 0)` rows have an upper bound below 0; those are the
harmed units. In the rows between, the interval contains 0, and the data
cannot say whether those effects are positive, negative, or zero.

## The teacher professional-development experiment

The `electric_teachers` data come from an experiment with 233 elementary
school teachers in 7 sites. Within each site, the researchers assigned
teachers at random to one of three versions of a professional development
program in science or to a control group. The variable `TxAny` is 1 for the
164 teachers assigned to any version of the program and 0 for the 69 in the
control group. The outcome `gain` is a teacher's score on a test of
knowledge about electric circuits after the program minus the score before
it, in percentage points.

Because the researchers randomized within sites, our analysis compares
teachers only with other teachers in their own site. We tell `cmrss()` the
sites by writing `| Site` in the formula. Computing bounds within sites
requires an optimization solver, such as the one in the `highs` package.
As in the simulation, we compute the helped and harmed counts at
`alpha = 0.025` each.

```{r teachers, eval = has_highs}
data(electric_teachers)
t_helped <- cmrss(gain ~ TxAny | Site, data = electric_teachers,
                  set = "all", alpha = 0.025, nperm = nperm, tol = tol,
                  opt.method = "ILP_highs")
t_harmed <- cmrss(-gain ~ TxAny | Site, data = electric_teachers,
                  set = "all", alpha = 0.025, nperm = nperm, tol = tol,
                  opt.method = "ILP_highs")
t_helped
t_harmed
t_n_helped <- sum(t_helped$bounds$lower > 0)
t_n_harmed <- sum(t_harmed$bounds$lower > 0)
```

```{r teachers-skip, include = FALSE}
# Without highs the chunk above does not run; the sentences below then
# report that instead of numbers.
if (!has_highs) t_n_helped <- t_n_harmed <- NA
```

Both calls warn that 29 of the 233 teachers share a single value of `gain`.
`cmrss()` ranks tied outcomes by their order in the data frame. With this
many ties, a different order of the rows could give somewhat different
bounds.

`r if (!has_highs) "These numbers were not computed, because the highs package is not installed." else paste0("With 95 percent confidence, at least ", t_n_helped, " teachers gained from the program.")`
`r if (has_highs && t_n_harmed == 0) "Analyzing \x60-gain\x60, we cannot show that any teacher was harmed. A failure to show harm is not evidence that no teacher was harmed." else if (has_highs) paste0("Analyzing \x60-gain\x60, we find at least ", t_n_harmed, " harmed teachers.") else ""`
`r if (has_highs) paste0("From the analysis of \x60gain\x60, as many as 233 - ", t_n_helped, " = ", 233 - t_n_helped, " teachers could have been harmed.") else ""`

The same plot for the teachers, analyzed within sites, shows for which
sorted effects the data rule out 0, and on which side.

```{r teachers-plot, eval = has_highs, fig.width = 6, fig.height = 5, fig.alt = "Lower and upper bounds for each of the 233 sorted effects of the professional development program on gain."}
plot_bounds(t_helped$bounds$lower, -rev(t_harmed$bounds$lower),
            xlab = "effect on gain, in percentage points")
```

`r if (has_highs) paste0("In the top ", t_n_helped, " rows the lower bound is above 0. ", if (t_n_harmed == 0) "No row has an upper bound below 0: the dashed line stays to the right of the vertical line at 0." else paste0("In the bottom ", t_n_harmed, " rows the upper bound is below 0.")) else ""`

## Choices that change the answer

With the argument `set`, we choose whose effects are bounded: `"treat"`,
`"control"`, or `"all"`.

With `alpha`, we choose the confidence level, 1 - `alpha`. When we report
the helped and harmed counts together, we compute each at `alpha / 2`.

With `s`, we choose the rank statistics that `cmrss()` combines. A rank
statistic replaces each outcome by its rank and adds up a score for the rank
of every treated unit. By default `cmrss()` uses polynomial scores: the
unit at rank r among n, counting from the smallest outcome, gets the score
$(r/(n + 1))^{s - 1}$. With s = 2 the score is proportional to the rank, so
every rank counts. With a large s nearly all the score sits on the few
largest outcomes. With a small s the combined test has more power against
effects shared by most units. With a large s it has more power against
large effects in a few units. The default combines three values of s, from
2 up to a maximum set by the numbers of treated and control units. In the
simulation above they were `r paste(helped$s, collapse = ", ")`. The section
"Choosing s" in `?cmrss` gives the rule, the simulations behind it, and the
alternative Stephenson scores (`scores = "stephenson"`).

The method is built for outcomes that take many different values. With a
binary outcome, every observed outcome is 0 or 1, so each unit is tied with
every other unit that has the same value. The order of the rows then
decides most of the ranks, and `cmrss()` warns. Every individual effect is
also $-1$, $0$, or $1$, so a bound can say little more than whether an
effect is at least 0 or at least 1. With a binary outcome there are four kinds of
unit: $Y_i(1) = Y_i(0) = 1$, $Y_i(1) = Y_i(0) = 0$, $Y_i(1) = 1$ and
$Y_i(0) = 0$ (helped), and $Y_i(1) = 0$ and $Y_i(0) = 1$ (harmed). Methods
built on the counts of these four kinds, such as those of Rigdon and
Hudgens (2015, *Statistics in Medicine* 34, doi:10.1002/sim.6384) and Li and
Ding (2016, *Statistics in Medicine* 35, doi:10.1002/sim.6924), give exact
confidence intervals for the average effect on a binary outcome.
